Lessons learned from bugs in models of human history

### 2020-06-02 paper.tex
Minor tweaks and add acks.

In the effort to build more realistic simulations of genetic diversity,
scientific software developers often focus on computational speed and biological realism.
Documentation and application programming interface (API), however, also determine
whether software is used, and whether it is used correctly.

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is its Python API, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ require users to specify cryptic command
line options to describe demographic models.
For example, the~\citet{gutenkunst2009inferring} demographic model
(that is the subject of this note), as written
in \ms\ syntax, is
\begin{minted}[]{tcsh}
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{minted}
\noindent This model is relatively simple,
and models with many more populations and parameters are increasingly
common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
The Python interface for \msprime, by contrast, allows the user to
state models in a more human-readable and programmatic manner,
and has many advantages over \ms's command line interface.

@@

\section*{Acknowledgements}
\aprcomment{to fill in}
\begin{itemize}
\item Alicia - sharing data, comments?
\item Ilan - comments
\item Chris for comments?
\end{itemize}

>>>>

In the effort to build more realistic simulations of genetic diversity,
scientific software developers often focus on computational speed and biological realism.
As the models simulated become more realistic, however, they also become
more complex and difficult to specify. The interface through which
users define their models is therefore increasingly important. Without
an intuitive and thoroughly documented interface, it is very difficult
to simulate complex population models correctly.

% Documentation and , however, also determine
% whether software is used, and whether it is used correctly.

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is its Python application programming interface (API), which greatly
increases the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ require users to specify cryptic command
line options to describe demographic models.
For example, the~\citet{gutenkunst2009inferring} demographic model
(that is the subject of this note), as written
in \ms\ syntax, is
\begin{minted}[]{tcsh}
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{minted}
\noindent This model is relatively simple,
and models with many more populations and parameters are increasingly
common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
The Python interface for \msprime, by contrast, allows the user to
state models in a more human-readable and programmatic manner,
and has many advantages over \ms's command line interface.

@@

\section*{Acknowledgements}
We would like to thank Alicia Martin for comments on the manuscript
and for providing the simulation data from~\cite{martin2017human}
used in Fig.~\ref{fig:prs}. We also like to thank
Chris Gignoux and Ilan Gronau for helpful comments on the manuscript.
JK would like to apologise
to those affected by the error in the example Out-of-Africa model and
by the poor design decisions in \msprime's demography API.

% JK: I left out the bit about thanking Alicia for the data - seems
% unnecessary?
% \aprcomment{to fill in}
% \begin{itemize}
% \item Alicia - sharing data, comments?
% \item Ilan - comments
% \item Chris for comments?
% \end{itemize}
